Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2016-01-14 docs/ms_sageAbundance.Rmd
adds year effect to sims; updates docs with Mevin's comments

Abstract
--------
Plant population models are powerful tools for predicting climate change impacts in one location, but are difficult to apply at landscape scales.
Here we overcome this limitation by taking advantage of two recent advances: remotely-sensed, species-specific estimates of plant cover and statistical models developed for spatio-temporal dynamics of animal populations.
Using computationally efficient model reparameterizations we fit a spatiotemporal population model to a 28 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming with rigorous accounting of spatial autocorrelation.
We include interannual variation in precipitation and temperature as covariates in the model to investigate how climate affects the cover of this species.
We then use the model to forecast the future abundance of sagebrush at the landscape scale under projected climate change, generating spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale. 
Our broad-scale and long-term predictions are rooted in small-scale and short-term population dynamics and provide an alternative to predictions offered by species distribution models that do not include population dynamics. 
Our approach is an important first step to using remote sensing data to model population responses to environmental change that play out at spatial scales far greater than the traditional field study plot.

@@

The ideal tool would be a large-scale, dynamic population model [@Schurr2012a; @Merow2014], but developing useful models at this scale has been limited by the availability of time series data at large spatial extents and statistical methods for fitting high-dimensional spatial models.
Fortunately, new advances in remote sensing and statistics now allow us to overcome both of these limitations.
First, new remote-sensing (RS) methods are now producing accurate time series of species-specific plant cover at landscape scales.  These data can be fit with dynamic population models which include yearly fluctuations in climate as covariates.
Such RS time series have revolutionized models of how climate affects ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
Second, animal population modelers have developed efficient dimension reduction and reparameterization techniques to efficiently fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
These new statistical methods have yet to be applied to RS-derived plant population data at broad scales.

Large-scale, spatially-explicit population models based on RS data could offer a valuable new way to investigate the effects of large-scale environmental changes playing out at landscape and regional scales.
Currently most assesments of how plant and animal populations will respond to climate change rely on species distribution models (SDMs).
SDMs rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009] and they are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014]. 
However, the short-term and small-scale population dynamics that actually drive the large-scale distributions of species are not represented in most SDMs. 
Since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [@Ehrlen2015]. 
Furthermore, because they lack short-term dynamics, SDMs usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached. Direct validation of such predictions is extremely rare [@Roberts2012].
Large-scale dynamic population models could overcome these limitations.
They would produce spatially-explicit estimates of species abundance within the species range [@Ehrlen2015] and would provide testable predictions of how populations should respond to short-term climate perturbations.
These short-term predictions also would give modelers the opportunity to repeatedly validate and refine their models [@Luo2011a].

Sagebrush (*Artemisia* spp.) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data. Sagebrush species are widely distributed [@Kuchler1964], they are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Harte2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015]. 
Ecohydrology models supply a possible mechanism for sagebrush losses predicted by SDMs: climate warming could lead to earlier snowmelt, increased evaporation and ultimately less recharge of deeper soil layers in the spring [@Schlaepfer2012; @Schlaepfer2014]. 
In warmer parts of its range, this could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014]. 
In contrast, at higher elevations and in colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].
A large-scale, spatially-explicit population model for sagebrush driven by interannual climate variability would provide a valuable new tool for assesing how sagebrush could respond to climate change in the future.

@@

Materials and Methods
---------------------
### Data
#### Remotely-sensed time series
To demonstrate our modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) canopy cover in Wyoming [@Homer2012].
As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65 and an out-of-sample RMSE of 5.46 [@Homer2012].
To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. 
This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011.
In this dataset, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Fig. 1).
Thus, the full dataset contains 369,603 observations spanning 27 year-to-year transitions (27 years $\times$ 13,689 pixels).

#### Climate covariates
Our approach is based on modeling how interannual changes in plant cover relate to seasonal climate variables.
We downloaded daily historic weather data for the center of our study site from the NASA Daymet data set (*available online*)[^link].
The Daymet weather data are interpolated between coarse observation units and capture some spatial variation.
We relied on weather data for the centroid of our study area.
We calculated five climate variables from the Daymet data for the time period coinciding with our remotely sensed data (1984 to 2011).

We narrowed our focus to climate covariates we know are important for sagebrush and that could be calculated from general circulation model projections.
The five climate variables in our population model are: (1) cumulative, "water year" precipitation for year \emph{t}-2 (*lagPpt*), (2) year \emph{t}-1 fall through summer precipitation (*ppt1*), (3) year \emph{t} fall through summer precipitation (*ppt2*), (4) year \emph{t}-1 average spring temperature (*TmeanSpr1*), and (5) year \emph{t} average spring temperature (*TmeanSpr2*), where *t-1* to *t* is the transition of interest.
We selected these variables a priori based on previous studies (see *Introduction*),  though not all emerge as important predictors in our model.
We could have applied a variety of model selection techniques [@Hooten2015], but chose not to because our focus is on the general approach.

@@

The spatial random effect ($\boldsymbol{\eta}$) accounts for spatial autocorrelation among pixels that occur near each other in space.
Thus, $\boldsymbol{\eta}$ acts as an offset on the intercept ($\beta_{0,t}$), creating a spatial field that defines how pixels differ from the mean, on average, in space (e.g., areas of perennially low or high cover, relative to average cover).
Fitting the model with a spatial random effect ($\eta$) is computationally demanding for large data sets like ours.
This is due to the required calculations of the spatial covariance matrices, which increase as a cubic function of the number of locations [@Wikle2010].
Key to our approach is a dimension reduction strategy that greatly reduces the number of parameters needed to be estimated to account for spatial variation by reducing the size of the spatial covariance matrices that need to be inverted at each MCMC iteration.
Fitting models that appropriately account for spatial autocorrelation over large spatial extents would not be feasible without these modern techniques.
Our dimension reduction strategy expresses the high dimensional spatial random effect, $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$ [e.g., @Hooten2003; @Hooten2007a; @Conn2015].
We can then define the spatial effect as

\vspace{-4em}
\begin{align}
\boldsymbol{\eta} &= \textbf{K}\boldsymbol{\alpha}, \\
\alpha_{j} &\sim \text{Normal}(0,\sigma_{\eta}^2).
\end{align}
\vspace{-4em}

@@

The last remaining obstacle is to parameterize \textbf{K}.
We use kernel convolution [@Barry1996; @Higdon1998] to interpolate the spatial random effect between \emph{m} "knots" that are nonrandomly distributed across the space of our study area.
This means we are modeling spatial random effects at the knot level, and we use \textbf{K} to interpolate those effects between knots.
We use an exponential kernel density to define the distance-decay function around the knots (\textbf{w}), such that the entries of \textbf{K} are

@@

and $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
However, given the relative size of our dataset and computational limitations, we defined kernels around 231 knots (Fig. B2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (~500 meters; Appendix B).
An infinite number of knots would result in a perfectly accurate representation of the spatial process and covariance model.
Since using an infinite number of knots is not possible, any use of the dimension reduction technique is an approximation, where the accuracy increases with the number of knots.
Given the tradeoff between knot number and computation time, we chose to base our knot number on the spatial dependence as described above.

@@

### Accomodating zeros
Our process model (Eq. 2) includes a log transformation of the observations ($\text{log}(y_{t-1}$)). 
Thus, our model does not accomodate zeros. 
Fortunately, we had very few instances where pixels went had 0% cover at time *t-1* (*N* = 47, which is 0.01% of the data set). 
Thus, we excluded those pixels from the model fitting process. 
However, when simulating the process, we needed to include possible transitions from zero to non-zero percent cover. 
We fit an intercept-only logistic model to estimate the probability of a pixel going from zero to non-zero cover

\vspace{-4em}
\begin{align}
\text{logit}(\mu_{i}) &= \alpha \\
y_{i} &\sim \text{Bernoulli}(\mu_{i})
\end{align}
\vspace{-4em}

where **y** is a vector of 0s and 1s corresponding to whether a pixel was colonized (>0% cover) or not (remains at 0% cover) and $\mu_i$ is the expected probability of colonization as a function of the mean probability of colonization ($\alpha$). 
We fit this simple model using the `glm' command in R [@R2013]. For data sets in which zeros are more common and the colonization process more important, the same spatial statistical approach we used for our cover change model could be applied.

### Fitting the model
We fit the model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1,000 iterations each, after discarding an initial 1,000 iterations as burn in.
Such short MCMC chains may seem unusual to those more familiar with other MCMC samplers (e.g., JAGS or WinBUGS).
However, the Stan algorithm is extremely efficient, meaning that, compared to other samplers, fewer iterations are required to achieve convergence.
Each MCMC chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
Model fitting required five days.
We assessed convergence visually and calculated scale reduction factors (Appendix C, $\hat{R}$ < 1.1 for all parameters) [@Gelman1992; @Gelman2009].

### Simulating the process
We performed four sets of simulations to (1) compare observed and simulated equilibrium cover, (2) compare observed and simulated year- and location-specific cover, (3) forecast future equilibrium population states under projected climate change, and (4) make temporally-explicit forecasts of sagebrush cover starting the final year of our observations and ending in year 2098. 
Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a perturbed climate time series (for 3).
We ran equilibrium simulations (1 and 3) for 2,000 time steps and then compared the output across simulations, after discarding an initial 100 time steps.
To calculate average future equilibrium sagebrush cover, we ran simulation (3) for each GCM and RCP scenario separately, and then averaged the results over GCMs.
For simulation (2), we initialized each pixel with its actual value for time *t* and cell *s* and projected the model forward one time step and compared the one-step ahead forecast with the observed value.
For simulation (4), we initialized each pixel with the final observed value in 2011 and then projected the model forward based on GCM yearly weather projections.
We ran these simulations for each GCM and RCP scenario combination separately and then aggregated the results over the GMCs by calculating the mean and the 90th percentiles for each RCP scenario.
We used the posterior mean of each parameter for all simulations except for (4) where we ran 50 simulations with unique sets of parameters from the MCMC chains.
The random year effects were not used in any simulation.

We required future projections of climate for our study area to conduct the equilibrium and temporally-explicit forecasts described above.
We used the most recent climate projections from the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
The CMIP5 provides projections from a suite of global circulations models (GCMs);  we used projections from 18 GCMs (Table A1) that produced weather projections for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and ongoing increase in greenhouse gas emissions, respectively.

@@

For the temporally-explicit forecasts we used yearly GCM projections from 2012 to 2098 to simulate the process starting from the end point of the remotely sensed sagebrush cover data (ends in 2011). 
We aggregated daily GCM output for each GCM and RCP scenario into the seasonal climate covariates used to fit our model.
These yearly climate time series were not aggregated further since we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.

@@

When we simulated the pixel-based population model based on observed climate it was able to reproduce the spatial pattern of observed percent cover, averaged over time (Fig. 4A,B).
Our model shows a tendency to underpredict perennially-low percent cover pixels (Fig. 4C), but does a better job at predicting high cover pixels.
Point predictions are most confident, though slightly biased, in low percent cover pixels (Fig. 4D).
The model is also able to adequately reproduce observed dynamics when we make one-step-ahead predictions based on observed climate and cover in the previous year for each pixel.
When we made these in-sample, one-step-ahead forecasts, the model achieved an RMSE = 4.31, in units of percent cover.
The correlation between observations and predictions was 0.62.

@@

Discussion
-----------
Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
Here, we developed a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
We first discuss our results specific to sagebrush ecology and response to climate, and then discuss the more general implications and limitations of our proposed approach.

### Sagebrush response to climate and climate change
The climate effects we estimated, based on cover data at 30 meter spatial resolution, are consistent with individual-level responses of sagebrush to climate-related variables.
Research on individual plants has shown that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and that warmer spring temperatures may enhance sagebrush growth in cold climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
In agreement with those individual-level responses, posterior means for all precipitation and temperature effects in our model were positive, except for the effect of fall-through-spring precipitation in the first year of a cover transition (*ppt1*, Fig. 3).
The cumulative amount of precipitation the year before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
However, the mean estimates for the climate effects are relatively weak (Fig. 3).
Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
Given findings from previous research demonstrating the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate.
Our model could be poorly specified since we used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.
However, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration [e.g., @Dalgleish2011; @Schlaepfer2014a].
More likely, aggregated estimates of plant abundance, such as percent cover, mask interannual variability at the level of the individual plant and makes it more difficult to detect the drivers of internanual variability. 
Nonetheless, our model was capabable of detecting climate effects, albeit small, that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.

>>>>

Abstract
--------
Plant population models are powerful tools for predicting climate change impacts in one location, but are difficult to apply at landscape scales.
We overcome this limitation by taking advantage of two recent advances: remotely-sensed, species-specific estimates of plant cover and statistical models developed for spatio-temporal dynamics of animal populations.
Using computationally efficient model reparameterizations, we fit a spatiotemporal population model to a 28 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming while formally accounting for spatial autocorrelation.
We include interannual variation in precipitation and temperature as covariates in the model to investigate how climate affects the cover of sagebrush.
We then use the model to forecast the future abundance of sagebrush at the landscape scale under projected climate change, generating spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale. 
Our broad-scale and long-term predictions are rooted in small-scale and short-term population dynamics and provide an alternative to predictions offered by species distribution models that do not include population dynamics. 
Our approach is among the first steps toward using remote sensing data to model population responses to environmental change that play out at spatial scales far greater than the traditional field study plot.

@@

The ideal tool would be a large-scale, dynamic population model [@Schurr2012a; @Merow2014], but developing useful models at this scale has been limited by the availability of time series data at large spatial extents and statistical methods for fitting high-dimensional spatial models.
Fortunately, new advances in remote sensing and statistics now allow us to overcome both of these limitations.
First, new remote-sensing (RS) methods are now producing accurate time series of species-specific plant cover at landscape scales.  These data can be fit with dynamic population models which include yearly fluctuations in climate as covariates.
Such RS time series have revolutionized models of how climate affects ecosystem-level processes [e.g., @Running2004] and have been used to detect long term trends in plant population abundance [e.g., @Homer2015], but they have yet to be used to drive a dynamic population model.
Second, animal population modelers have developed dimension reduction and reparameterization techniques to efficiently fit high-dimension spatiotemporal models [see @Conn2015 for a review]. 
These new statistical methods have yet to be applied to RS-derived plant population data at broad scales.

Large-scale, spatially-explicit population models based on RS data could offer a valuable new way to investigate the effects of large-scale environmental changes playing out at landscape and regional scales.
Most current assesments of how plant and animal populations will respond to climate change rely on species distribution models (SDMs).
SDMs rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009] and they are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014]. 
However, the short-term and small-scale population dynamics that actually drive the large-scale distributions of species are not represented in most SDMs. 
Because SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [@Ehrlen2015]. 
Furthermore, because they lack short-term dynamics, SDMs usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached. Direct validation of such predictions is extremely rare [@Roberts2012].
Large-scale dynamic population models could overcome these limitations.
They would produce spatially-explicit estimates of species abundance within the species range [@Ehrlen2015] and would provide testable predictions of how populations should respond to short-term climate perturbations.
These short-term predictions also would give modelers the opportunity to repeatedly validate and refine their models [@Luo2011a].

Sagebrush (*Artemisia* spp.) ecosystems offer an ideal testing ground for new spatially explicit population models derived from RS data. Sagebrush species are widely distributed [@Kuchler1964], they are sensitive to climate [@Dalgleish2011; @Schlaepfer2014; @Schlaepfer2014a; @Miglia2005; @Poore2009; @Apodaca2013; @Perfors2003; @Harte2015], new landscape and regional scale time series of sagebrush cover are now being produced from aerial imagery [@Homer2012], and forecasts of future sagebrush ecosystems are in high demand due to the precarious conservation status of the greater sage-grouse (*Centrocercus urophasianus*) [@Arnett2015].
SDMs typically predict that much of the area occupied by sagebrush ecosystems today will become unsuitable for sagebrush due to climate change, resulting in a dramatic loss in the extent of sagebrush habitat by the end of this century [@Shafer2001; @Neilson2005; @Bradley2010; @Schlaepfer2012; @Still2015]. 
Ecohydrology models supply a possible mechanism for sagebrush losses predicted by SDMs: climate warming could lead to earlier snowmelt, increased evaporation and ultimately less recharge of deeper soil layers in the spring [@Schlaepfer2012; @Schlaepfer2014]. 
In warmer parts of its range, increased temperature could be especially detrimental to sagebrush as it depends on water from deeper soil to survive and grow in this arid region [@Pechanec1937; @Schlaepfer2011; @Germino2014]. 
In contrast, at higher elevations and in colder regions, warming and earlier snowmelt could lengthen the growing season and increase sagebrush occurrence [@Schlaepfer2012; @Schlaepfer2014].
Direct observations of individual plants and experimental plots tend to agree with these models:  growth tends to respond negatively to spring and summer temperatures [@Miglia2005; @Poore2009; @Apodaca2013] except at higher elevations where earlier snowmelt may allow for a longer growing season [@Perfors2003; @Harte2015].
A large-scale, spatially-explicit population model for sagebrush driven by interannual climate variability would provide a valuable new tool for assessing how sagebrush could respond to climate change in the future.

@@

Materials and Methods
---------------------
### Data
#### Remotely-sensed time series
To demonstrate our modeling approach, we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) canopy cover in Wyoming [@Homer2012].
As part of a separate study, sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming.
The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65 and an out-of-sample RMSE of 5.46 [@Homer2012].
To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. 
This resulted in an annual time series of sagebrush cover at 30 meter resolution from 1984 to 2011.
In this dataset, values represent the percentage of a 30 $\times$ 30 meter pixel covered by sagebrush.
In our study, we focused on a 5,070 $\times$ 2,430 meter subset totaling 13,689 30 $\times$ 30 meter pixels each year (Fig. 1).
Thus, the full dataset contains 369,603 observations spanning 27 year-to-year transitions (27 years $\times$ 13,689 pixels).

#### Climate covariates
Our approach models interannual changes in plant cover as a function of seasonal climate variables.
We used daily historic weather data for the center of our study site from the NASA Daymet data set (*available online*)[^link].
The Daymet weather data are interpolated between coarse observation units and capture some spatial variation.
We relied on weather data for the centroid of our study area.
We calculated five climate variables from the Daymet data for the time period coinciding with our remotely sensed data (1984 to 2011).

We narrowed our focus to climate covariates we know are important for sagebrush and that could be calculated from general circulation model projections.
The five climate variables in our population model are: (1) cumulative, "water year" precipitation for year \emph{t}-2 (*lagPpt*), (2) year \emph{t}-1 fall through summer precipitation (*ppt1*), (3) year \emph{t} fall through summer precipitation (*ppt2*), (4) year \emph{t}-1 average spring temperature (*TmeanSpr1*), and (5) year \emph{t} average spring temperature (*TmeanSpr2*), where *t-1* to *t* is the transition of interest.
We selected these variables *a priori* based on previous studies (see *Introduction*),  though not all emerge as important predictors in our model.

@@

The spatial random effect ($\boldsymbol{\eta}$) accounts for spatial autocorrelation among pixels that occur near each other in space.
Thus, $\boldsymbol{\eta}$ acts as an offset on the intercept ($\beta_{0,t}$), creating a spatial field that defines how pixels differ from the mean, on average, in space (e.g., areas of perennially low or high cover, relative to average cover).
Fitting the model with a spatial random effect ($\eta$) is computationally demanding for large data sets like ours.
The computational demand is due to the required calculations of the spatial covariance matrices, which increase as a cubic function of the number of locations [@Wikle2010].
Key to our approach is a dimension reduction strategy that greatly reduces the number of parameters needed to be estimated to account for spatial variation by reducing the size of the spatial covariance matrices that need to be inverted at each MCMC iteration.
Fitting models that appropriately account for spatial autocorrelation over large spatial extents would not be feasible without these modern techniques.
Our dimension reduction strategy expresses the high dimensional spatial random effect, $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$ [e.g., @Hooten2003; @Hooten2007a; @Conn2015].
We can then approximate the spatial effect as

\vspace{-4em}
\begin{align}
\boldsymbol{\eta} &\approx \textbf{K}\boldsymbol{\alpha}, \\
\alpha_{j} &\sim \text{Normal}(0,\sigma_{\eta}^2).
\end{align}
\vspace{-4em}

@@

The last remaining obstacle is to parameterize the matrix of basis functions, \textbf{K}.
We use kernel convolution [@Barry1996; @Higdon1998] to interpolate the spatial random effect between \emph{m} "knots" that are nonrandomly distributed across the space of our study area.
This means we are modeling spatial random effects at the knot level, and we use \textbf{K} to interpolate those effects between knots.
We use an exponential kernel density to define the distance-decay function around the knots (\textbf{w}), such that the entries of \textbf{K} are

@@

and $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth.
It is possible, through exhaustive model selection and fitting, to determine the optimal form of the kernel and to estimate optimal values for $\sigma$ [@Higdon2002; @Hooten2015].
However, given the relative size of our dataset and computational limitations, we defined kernels around 231 knots (Fig. B2) whose nearest neighbor distances are approximately equal to the range of spatial dependence in residuals from a simple GLM fit without climate covariates and the spatial random effect (~500 meters; Appendix B).
An infinite number of knots would result in an exact representation of the spatial process and covariance model.
Computationally, using an infinite number of knots is not possible, thus the use of dimension reduction techniques serves as an approximation, where the accuracy increases with the number of knots.
Given the tradeoff between knot number and computation time, we chose to base our knot number on the spatial dependence as described above.

@@

### Accomodating zeros
Our process model (in Eq. 2) includes a log transformation of the observations ($\text{log}(y_{t-1}$)). 
Thus, our model does not accomodate zeros. 
Fortunately, we had very few instances where pixels had 0% cover at time *t-1* (*N* = 47, which is 0.01% of the data set). 
Thus, we excluded those pixels from the model fitting process. 
However, when simulating the process, we needed to include possible transitions from zero to non-zero percent cover. 
We fit an intercept-only logistic model to estimate the probability of a pixel going from zero to non-zero cover

\vspace{-4em}
\begin{align}
y_{i} &\sim \text{Bernoulli}(\mu_{i}) \\
\text{logit}(\mu_{i}) &= b_0
\end{align}
\vspace{-4em}

where **y** is a vector of 0s and 1s corresponding to whether a pixel was colonized (>0% cover) or not (remains at 0% cover) and $\mu_i$ is the expected probability of colonization as a function of the mean probability of colonization ($b_0$). 
We fit this simple model using the `glm' command in R [@R2013]. For data sets in which zeros are more common and the colonization process more important, the same spatial statistical approach we used for our cover change model could be applied.

### Fitting the model
We fit the spatiotemporal model in R [@R2013] using the 'No-U-Turn' Hamiltonian Monte Carlo sampler in Stan [@stan2014] and the RStan package [@rstan2014].
We obtained posterior distributions of all model parmaters from three MCMC chains comprised of 1,000 iterations each, after discarding an initial 1,000 iterations as burn in.
Short chains of samples are a hallmark of the Stan algorithm, which is extremely efficient.
Compared to other samplers, fewer iterations are required to achieve convergence.
Each chain was initialized with unique parameter values and the model was fit in parallel using the Utah State University High-Performance Computing facility.
Model fitting required five days.
We assessed convergence visually and calculated scale reduction factors (Appendix C, $\hat{R}$ < 1.1 for all parameters) [@Gelman1992; @Gelman2009].

### Simulating the process
We performed four sets of simulations to (1) compare observed and simulated equilibrium cover, (2) compare observed and simulated year- and location-specific cover, (3) forecast future equilibrium population states under projected climate change, and (4) make temporally-explicit forecasts of sagebrush cover starting the final year of our observations and ending in year 2098. 
Using the posterior distribution of model parameters, we simulated a matrix of pixels equal to the size of the study area (13,689 pixels or matrix elements).
For simulations (1) and (3) we initialized all pixels with arbitrarily low cover (1%) and then projected the model forward by randomly drawing climate covariates from the observed climate time series (for 1) or a perturbed climate time series (for 3).
We ran equilibrium simulations (1 and 3) for 2,000 time steps and then compared the output across simulations, after discarding an initial 100 time steps.
To calculate average future equilibrium sagebrush cover, we ran simulation (3) for each GCM and RCP scenario separately, and then averaged the results over GCMs.
For simulation (2), we initialized each pixel with its actual percent cover value for time *t* and cell *s* and projected the model forward one time step and compared the one-step ahead forecast with the observed value.
For simulation (4), we initialized each pixel with the final observed value in 2011 and then projected the model forward based on GCM yearly weather projections.
We ran these simulations for each GCM and RCP scenario combination separately and then aggregated the results over the GMCs by calculating the mean and the 90th percentiles for each RCP scenario.
We used the posterior mean of each parameter for all simulations except for (4) where we ran 50 simulations with unique sets of parameters from the chains.
The random year effects were not used in simulations 1-3 where.

We required future projections of climate for our study area to conduct the equilibrium and temporally-explicit forecasts described above.
Thus, we used the most recent climate projections from the Intergovernmental Panel on Climate Change (IPCC), the Coupled Model Intercomparison Project 5 (CMIP5; *available online*)[^link2].
The CMIP5 provides projections from a suite of global circulations models (GCMs);  we used projections from 18 GCMs (Table A1) that produced weather projections for three ``Representative Concentration Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5 (*described online*)[^link3].
The three RCPs correspond to stabilization of radiative forcing before 2100, after 2100, and ongoing increase in greenhouse gas emissions, respectively.

@@

For the temporally-explicit forecasts we used yearly GCM projections from 2012 to 2098 to simulate the process starting from the end point of the remotely sensed sagebrush cover data (ends in 2011). 
We aggregated daily GCM output for each GCM and RCP scenario into the seasonal climate covariates used to fit our model.
These yearly climate time series were not aggregated further because we ran simulations for each GCM and RCP scenario, rather than one simulation per RCP scenario averaged over GCMs.

@@

When we simulated the pixel-based population model based on observed climate, it was able to reproduce the spatial pattern of observed percent cover, averaged over time (Fig. 4A,B).
Our model shows a tendency to underpredict perennially-low percent cover pixels (Fig. 4C), but does a better job at predicting high cover pixels.
Point predictions are most confident, though slightly biased, in low percent cover pixels (Fig. 4D).
The model is also able to adequately reproduce observed dynamics when we make one-step-ahead predictions based on observed climate and cover in the previous year for each pixel.
When we made these in-sample, one-step-ahead forecasts, the model achieved an RMSE = 4.31, in units of percent cover.
The correlation between observations and predictions was 0.62.

@@

Discussion
-----------
Despite the need to forecast population responses to climate change over large spatial extents, as demonstrated by the wide application of species distribution models [e.g., @Clark2014], landscape-scale population models for plant species remain more concept than reality [@Schurr2012a; @Merow2014].
We developed a new approach using remotely-sensed time series of plant percent cover [@Homer2012] and methods from the spatio-temporal modeling of animal abundance [e.g., @Conn2015].
We first discuss our results specific to sagebrush ecology and response to climate, and then discuss the more general implications and limitations of our proposed approach.

### Sagebrush response to climate and climate change
The climate effects we estimated, based on cover data at 30 meter spatial resolution, are consistent with individual-level responses of sagebrush to climate-related variables.
Research on individual plants has shown that wetter winters are correlated with greater stem growth in sagebrush [@Poore2009; @Apodaca2013] and that warmer spring temperatures may enhance sagebrush growth in cold climates by advancing the date of snowmelt and increasing the length of the growing season [@Perfors2003; @Harte2015].
In agreement with those individual-level responses, posterior means for all precipitation and temperature effects in our model were positive, except for the effect of fall-through-spring precipitation in the first year of a cover transition (*ppt1*, Fig. 3).
The cumulative amount of precipitation the year before a cover transition (*pptLag* in our model) emerged as the strongest predictor of sagebrush cover change (Fig. 3).
However, mean estimates for the climate effects are relatively weak (Fig. 3).
Such small effects could indicate that sagebrush are not very sensitive to interannual climate variability, that our model is poorly specified, or that climate responses are difficult to detect using coarse-scale data.
Given findings from previous research demonstrating the importance of precipitation and temperature on sagebrush growth [@Pechanec1937; @Schlaepfer2011; @Germino2014] and regeneration [@Schlaepfer2014a], it is unlikely that sagebrush are insensitive to climate.
We used aggregated climate covariates that may not completely capture the climate-dependence of sagebrush cover change.
However, the covariates we chose closely match the climate-related variables that have been shown to drive sagebrush growth, survival, and regeneration [e.g., @Dalgleish2011; @Schlaepfer2014a].
More likely, aggregated estimates of plant abundance, such as percent cover, mask interannual variability at the level of the individual plant and makes it more difficult to detect the drivers of internanual variability. 
Nonetheless, our model was capabable of detecting climate effects that agree with our knowledge of sagebrush ecology and allowed us to make forecasts of future sagebrush abundance.
